\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{hyperref}

\title{Guide to the Current Brenier Formalization}
\author{}
\date{}

\begin{document}
\maketitle

\section*{Scope}
This note describes the architecture that is currently formalized in the repository
\texttt{OptimalTransport}, after the migration to proper extended-real convex potentials.

\section*{Final endpoint now formalized}
In finite-dimensional real inner-product spaces, for quadratic cost
\[
  c_\kappa(x,y)=\kappa\|x-y\|^2,\qquad \kappa>0,
\]
the repository now proves a classical Brenier-type endpoint with no extra Rockafellar boundedness
assumption:

\begin{itemize}
\item from optimality of \(\pi\), one obtains pairing cyclical monotonicity of
  \(\operatorname{supp}\pi\);
\item from that, one builds a proper convex potential \(\Phi:E\to\mathbb R\cup\{+\infty\}\),
  with
  \[
    \operatorname{supp}\pi \subseteq \operatorname{Graph}(\partial \Phi);
  \]
\item if \(\mu\ll m\) (Haar/Lebesgue), then
  \[
    \pi = (\mathrm{id}, \nabla(\Phi^\ast))_{\#}\mu,
    \qquad
    (\nabla(\Phi^\ast))_{\#}\mu=\nu,
  \]
  where in code \(\Phi^\ast\) is represented by \(x\mapsto (\Phi x).\texttt{untopD }0\) on the
  \(\mu\)-relevant locus.
\end{itemize}

The main assembled theorem is in
\texttt{OptimalTransport/Brenier.lean}:
\[
  \texttt{Coupling.exists\_properConvex\_potential\_of\_optimal\_quadratic}.
\]

\section*{File map}

\subsection*{1. Couplings and costs}
\begin{itemize}
\item \texttt{OptimalTransport/Coupling.lean}: transport plans, marginals, couplings, graph plans.
\item \texttt{OptimalTransport/Cost.lean}: \texttt{transportCostENNReal}, \texttt{transportCost},
  finite-second-moment helpers, quadratic cost.
\end{itemize}

\subsection*{2. Existence of optimal couplings}
\begin{itemize}
\item \texttt{OptimalTransport/Existence.lean}: existence for lower-semicontinuous
  \(\mathbb R_{\ge0}^{\infty}\)-valued costs on Polish spaces (Prokhorov + Portmanteau route).
\end{itemize}

\subsection*{3. Cyclical monotonicity from optimality}
\begin{itemize}
\item \texttt{OptimalTransport/CyclicMonotone.lean}:
  \(\texttt{CCyclicallyMonotone}\), local rectangle lemmas, finite perturbation competitor,
  and the global theorem
  \[
    \texttt{CCyclicallyMonotone\_support\_of\_optimal\_of\_integrable}.
  \]
\item Quadratic specialization:
  \[
    \texttt{pairing\_support\_of\_optimal\_quadratic}.
  \]
\end{itemize}

\subsection*{4. Proper convex layer}
\begin{itemize}
\item \texttt{OptimalTransport/ProperConvex.lean}:
  \texttt{ProperSubgradient}, \texttt{ProperSubgradientGraph}, \texttt{EffectiveDomain},
  and bridge lemmas to real-valued wrappers when needed.
\end{itemize}

\subsection*{5. Rockafellar in \(\mathbb R\cup\{+\infty\}\)}
\begin{itemize}
\item \texttt{OptimalTransport/Rockafellar.lean}:
  \texttt{properRockafellarPotential}, lower semicontinuity/measurability,
  pointwise finiteness on cyclically monotone supports, and global containment
  \[
    \texttt{subset\_ProperSubgradientGraph\_properRockafellarPotential\_of\_pairingClosedChainMonotone}.
  \]
\item Interface theorem from finite-cycle inequalities to list-based closed chains:
  \[
    \texttt{pairingClosedChainMonotone\_of\_cCyclicallyMonotone\_negInner}.
  \]
\end{itemize}

\subsection*{6. Brenier bridge}
\begin{itemize}
\item \texttt{OptimalTransport/Brenier.lean}:
  graph-collapse lemmas in the proper-convex setting and the final assembly theorem
  \[
    \texttt{Coupling.exists\_properConvex\_potential\_of\_optimal\_quadratic}.
  \]
\end{itemize}

\section*{Logical chain in the code}
The assembled proof now follows:
\[
\text{optimal quadratic coupling}
\Rightarrow
\text{pairing cyclical monotonicity of }\operatorname{supp}\pi
\Rightarrow
\text{proper Rockafellar potential }\Phi
\Rightarrow
\operatorname{supp}\pi\subseteq \operatorname{Graph}(\partial\Phi)
\Rightarrow
\text{graph-plan concentration}
\Rightarrow
\text{Monge map representation}.
\]

\section*{Status}
The former \texttt{HasRockafellarBound} bottleneck is removed from the active Brenier theorem path.
Compatibility lemmas with stronger assumptions are retained only as auxiliary corollaries.

\end{document}

